---
title: "UFO Programming Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using UFO Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

UFOs are primarily a library that aides you in the creation of custom R vector
backend. In this vignette we will show you how to implement a simple
larger-than-memory vector package by example. We will explain both how to write
the package from scratch and how to make UFO work.

Implementing a custom UFO vector requires you to write five pieces of code in R
and C:

- [R] constructors for your vectors
- [C] a population function
- [C] a destructor function
- [C] a structure that configures the population function
- [C] glue that feeds the previous two things to the UFO framework

# Example: UFO sequences

As an example, we will use an implementation of sequences. Our sequence will be
an integer vector that has a beginning, a step, and an end. Any given element
of the vector is equal to the previous element plus the step. You probably
already know them from R:

``` {r}
seq(from = 1, to = 10, by = 2)
```

Our sequences will be created by the following function:

```R
ufo_seq(from = 1, to = 10, by = 2)
```

And we're going to create a package around it called `ufoseq`.

## Creating a package

To create a rudimentary R package we create a directory called `ufoseq`
and inside this directory we create two subdirectories: `R` and `src`.

Then, we add a `DESCRIPTION` file to the `ufoseq` directory and fill it after
the following fashion:

```
Package: ufoseq
Type: Package
Title: Implementation of sequences using UFOs.
Description: Example implementation of UFO vectors that provides larger-than-memory sequence vectors.
Version: 1.0
Authors@R: c(person(given = "Konrad",  family = "Siek", role = c("aut", "cre"),
                    email = "siekkonr@fit.cvut.cz"))
Maintainer: Konrad Siek <siekkonr@fit.cvut.cz>
License: GPL-2 | GPL-3
Encoding: UTF-8
LazyData: true
Depends: ufos
LinkingTo: ufos
NeedsCompilation: yes
Suggests: 
    knitr,
    rmarkdown
VignetteBuilder: knitr
```

Note that we are adding the `ufos` package as both a dependency and a linking
requirement. We are doing this, because we will later on import some C
functions from `ufos`.

## R constructors

First, we create an R constructor for our vectors. Create an R script at
`ufoseq/R/ufoseq.R`. Here we essentially write a simple R function that just
calls a C function we will write later.

``` R
ufo_seq <- function(from, to, by = 1) {
  # Call the C function that actually creates the vector
  .Call("ufo_seq", from, to, by)
}
```

That's simple enough, but function should check whether the arguments it
receives are what they are expected to be. Thus, we need to add some simple
checks.

``` R
ufo_seq <- function(from, to, by = 1) {
  # check if any of the arguments were missing
  if (missing(from)) stop ("'from' is a required argument")
  if (missing(to)) stop ("'to' is a required argument")
      
  # check whether the arguments are non-zero length
  if (length(from) == 0) stop("'from' cannot be zero length")
  if (length(to) == 0) stop("'to' cannot be zero length")
  if (length(by) == 0) stop("'by' cannot be zero length")

  # check whether this sequence makes sense.
  if (from >= to) stop("'from' must not be less than 'to'")
  if (by <= 0) stop("'by' must be larger than zero")
  
  # check whether the arguments are of scalars
  if (length(from) > 1) 
    warn("'from' has multiple values, only the first value will be used")
  if (length(to) > 1) 
    warn("'to' has multiple values, only the first value will be used")
  if (length(by) > 1)
    warn("'by' has multiple values, only the first value will be used")
    
  # Convert inputs to integers and call the C function that actually creates
  # the vector
  .Call("ufo_seq", as.integer(from), as.integer(to), as.integer(by))
}
```

Now the function looks like it means business! ;) We then need to export this
function to the package namespace. Create a file `ufoseq/NAMESPACE` and fill it
out as follows:

```R
useDynLib(ufoseq, .registration = TRUE, .fixes = "")
export(ufo_seq)
```

## Some glue

Now onto the C function. We create a header file: `ufoseq/src/ufoseq.h` and
declare a the `ufo_seq` C function:

```C
#pragma once
#include "Rinternals.h"

SEXP ufo_seq(SEXP from, SEXP to, SEXP by);
```

This function takes three arguments of type SEXP and returns a SEXP. SEXPs are
a supertype of all R objects. They are declared in `Rinternals.h`, which is why
we must include it. In our case, to, from, and by are going to be integers or
doubles, and we will return a vector that is also an integer vector or a
double. It sometimes saves confusion to mark that in the function signatures
with comments:

```C
#pragma once
#include "Rinternals.h"

SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by);
```

Then, let us register the function with the R interpreter. We create a C file
`ufoseq/src/init.c`:

```C
#include "ufoseq.h"

#include <R_ext/Rdynload.h>
#include <R_ext/Visibility.h>

// List of functions provided by the package.
static const R_CallMethodDef CallEntries[] = {
    // Constructors
    {"ufo_seq",  (DL_FUNC) &ufo_seq,  3},

    // Terminates the function list. Necessary, do not remove.
    {NULL, NULL, 0}
};
```

Here we start by including the header file we previously created, so that we
can refer to our `ufo_seq` function. We also have some helpful R includes. More
important, below, we construct a struct that serves as the registry of C
functions that can be called from R. The function description consists of three
fields: a name, a pointer to a C function, and the number of arguments. In our
case the name is `ufo_seq`, we get the pointer from the reference to the
`ufo_seq` function defined in `ufoseq.h`, and the number of arguments is 3.

If we had more functions we would add them to this list. It's important to
terminate the list with a `{NULL, NULL, 0}` entry. Otherwise terrible things
happen.

Now we can start working on the C function that creates our UFO vectors. We add
a file `ufoseq/src/ufoseq.c` and start implementing our `ufoseq` function:

```C
#include "ufoseq.h"

SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by) {

    return R_NilValue;
}
```

We include the header where our function is declared. We then define the
function. For now it's empty and it returns R's `NULL` object. Before we start
filling it out we need to do some more things though. 

## UFO includes

First, we need to import some UFO definitions:

We need to create a directory `ufoseq/include`. We  need to copy the `ufos.h`
header file from the `ufos` package there. Then we need to create another
directory: `../include/mappedMemory/` and copy another header file,
`userfaultCore.h` there. These two files contain definitions of all the things
you need to work with UFOs:

  - `ufo_vector_type_t` - UFO vector type definitions, which are analogous to R
    vector types, and `ufo_type_to_vector_type`, a function to convert from one
    to the other,
  - `ufo_source_t` - a structure that passes the necessary configuration data
    to the UFO framework,
  - `ufo_initialize` and `ufo_shutdown` - functions for starting and shutting
    down the UFO framework,
  - `ufo_new` a generic constructor for UFO vectors.

## Creating a UFO

In order to create a UFO we use `ufo_new`, which is defined like this: 

```C
SEXP ufo_new(ufo_source_t*);
```

But we can't just use it, since it's defined in the `ufos` package. Instead, we
need to import it from the package. We do it like this:

```C
ufo_new_t ufo_new = (ufo_new_t) R_GetCCallable("ufos", "ufo_new");
```

Now we need to construct a structure of type `ufo_source_t` to pass as an
argument.

This structure is defined as follows:

```C
typedef struct {
    ufUserData*         data;
    ufPopulateRange     population_function;
    ufo_destructor_t    destructor_function;
    ufo_vector_type_t   vector_type;
    size_t              vector_size;
    size_t              element_size;
    int                 *dimensions;
    size_t              dimensions_length;
    int32_t             min_load_count;
} ufo_source_t;
```

Most of these are straightforward. `vector_type` is one of the following vector
types:

```C
typedef enum {
    UFO_CHAR = CHARSXP,
    UFO_LGL  = LGLSXP,
    UFO_INT  = INTSXP,
    UFO_REAL = REALSXP,
    UFO_CPLX = CPLXSXP,
    UFO_RAW  = RAWSXP
} ufo_vector_type_t;
```

Let us leave `data`, `population_function` and `destructor_function` for later. 

`vector_size` is the number of elements in the vector and `element_size` is the
size of each element in bytes. 

`dimensions` and `dimensions_length` are used to provide extra data for
matrices. If you are not writing a matrix, set `dimensions` to `NULL`.  If you
are writing a matrix, `dimensions_length` is the number of dimensions, and
`dimensions` will represent the sizes of each dimension.

Next, `min_load_count` reperesents the minimum number of elements to populate
when the vector is accessed. We will explain the details later, when writing
the population function. A good typical number is 1 megabyte's worth of
elements.

How do we fill these values? Most of them are simple, so can thus create a
`source` like this:

```C
SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by) {

    ufo_source_t* source = (ufo_source_t*) malloc(sizeof(ufo_source_t));

    source->vector_type = UFO_INT; 
    source->element_size = sizeof(int);

    int from_value = INTEGER_ELT(from, 0);
    int to_value = INTEGER_ELT(to, 0);
    int by_value = INTEGER_ELT(by, 0);
    source->vector_size = (to_value - from_value) / by_value
                        + ((to_value - from_value) % by_value > 0)

    source->dimensions = NULL;
    source->dimensions_length = 0;

    source->min_load_count = 1000000 / sizeof(int);

    ufo_new_t ufo_new = (ufo_new_t) R_GetCCallable("ufos", "ufo_new");
    return ufo_new(source);
}
```

We calculate the size by first calculating the division of the difference
between `to_value` and `from_value` and, then dividing the result by
`by_value`, and finally we take the ceiling of the division. Except we do it
more C-like.

## Application-specific data

The data in `source` are for the UFO framework. But we will also need some data
for our population function. That data goes into the `data` field. The UFO
framework does not care what you put there, this is only for you. That is why
the type of `data` is actually `void *`. You need to define what data you are
going to need yourself. So let us define the following structure in `ufoseq.h`

```C
    typedef struct {
        int from;
        int to;
        int by; 
    } ufo_seq_data_t;
```

We then initialize this structure and include it in `source`:

```C
SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by) {

    ufo_source_t* source = (ufo_source_t*) malloc(sizeof(ufo_source_t));

    source->vector_type = UFO_INT; 
    source->element_size = sizeof(int);

    int from_value = INTEGER_ELT(from, 0);
    int to_value = INTEGER_ELT(to, 0);
    int by_value = INTEGER_ELT(by, 0);
    source->vector_size = (to_value - from_value) / by_value
                        + ((to_value - from_value) % by_value > 0)

    source->dimensions = NULL;
    source->dimensions_length = 0;

    source->min_load_count = 1000000 / sizeof(int);

    ufo_seq_data_t *data = (ufo_seq_data_t*) malloc(sizeof(ufo_seq_data_t));
    data->from = from_value;
    data->to = to_value;
    data->by = by_value;
    source->data = (ufUserData) data;

    ufo_new_t ufo_new = (ufo_new_t) R_GetCCallable("ufos", "ufo_new");
    return ufo_new(source);
}
```

## Destructor function

We allocate some memory, we should clean up after ourselves. Generally
speaking, R has a garbage collector which will figure out when our vector stops
being in use. When this happens, the framework will try to clean up the various
objects it allocated. One of the first steps of that process is to call the
destructor function defined in the `source` struct. Inside this function it is
the programmer's job, ie. your job, to clean up your `data`.

This function has to have the following type:

```C
typedef void (*ufo_destructor_t)(ufUserData*)
```

This defines a function with one argument of type `ufUserData *`. This argument
is actually the structure we specified in `source->data` above. This means we
can just cast it to `ufo_seq_data_t*`, In our case this structure is
straightforward, s we can just deallocate it using `free`. So our destructor
looks like this:

```C
void destroy_data(ufUserData *data) {
    ufo_seq_data_t *ufo_seq_data = (ufo_seq_data_t*) data;
    free(ufo_seq_data);
}
```

We then attach this function to the `source` structure like so:

```C
SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by) {

    ufo_source_t* source = (ufo_source_t*) malloc(sizeof(ufo_source_t));

    source->vector_type = UFO_INT; 
    source->element_size = sizeof(int);

    int from_value = INTEGER_ELT(from, 0);
    int to_value = INTEGER_ELT(to, 0);
    int by_value = INTEGER_ELT(by, 0);
    source->vector_size = (to_value - from_value) / by_value
                        + ((to_value - from_value) % by_value > 0);

    source->dimensions = NULL;
    source->dimensions_length = 0;

    source->min_load_count = 1000000 / sizeof(int);

    ufo_seq_data_t *data = (ufo_seq_data_t*) malloc(sizeof(ufo_seq_data_t));
    data->from = from_value;
    data->to = to_value;
    data->by = by_value;
    source->data = (ufUserData) data;

    source->destructor_function = &destroy_data;

    ufo_new_t ufo_new = (ufo_new_t) R_GetCCallable("ufos", "ufo_new");
    return ufo_new(source);
}
```

## Population function

The UFO framework will call your population function whenever a new chunk of
memory in a vector is accessed. Therefore it is going to be the main thing that
defines what your custom vector does.  The type of this function is defined in
`userfaultCore.h`:

```C
typedef int (*ufPopulateRange)(uint64_t startValueIdx, uint64_t endValueIdx, ufPopulateCallout callout, ufUserData src, char* target);
```

So it's a function that takes a lot of arguments and returns an integer value.
The return value is supposed to be 0 if the function completes succesfully, and
any other value in case of an error. Let's take a look at the arguments.

First `target` is a pointer to an area of memory which you must fill with data.
This is where we will be writing our sequence.

Arguments `startValueIdx` and `endValueIdx` tell you which values you need to
generate during this access. `startValueIdx` is the first value and
`endValueIdx` is the exclusive limit. For instance, if somebody accesses your
vector in R as `v[1:100]` then `startValueIdx` will be 0 and `endValueIdx` will
be 100. This means you are supposed to fill in `((int *) target)[0]` all the
way through to `((int *) target)[99]`, but **not** `((int *) target)[100]`.

Note that C is 0-indexed and R is 1-indexed.

The area of memory pointed to by `target` is already appropriately offset. This
means that if somebody accesses your vector in R as `v[101:200]`, then
`startValueIdx` will be 100 and  `endValueIdx` will be 200, but you are
supposed to fill in `((int *) target)[0]` all the way through to `((int *)
target)[99]`. And the value you write to `((int *) target)[0]` should be the
value you would expect to see at `v[101]` in R.

It is also important to point out, that for the sake of efficiency, the UFO
framework will actually round up the amount elements that need to be generated
to the nearest memory page larger than the memory required to allocate
`source->min_load_count` elements.

Another important argument is `userData`. This is going to be the structure of
type `ufo_seq_data_t` that you passed in to `ufo_new` via `source`. This means it contains all
the necessary data for your vector to generate data.

The function `callout` is advanced functionality that lets you inform the UFO
framework that you would like to load more data at once, but we will not
explore this functionality here.

Let us instead write out population function for sequences.

```C
int populate(uint64_t startValueIdx, uint64_t endValueIdx,
             ufPopulateCallout callout, ufUserData userData, char* target) {

    ufo_seq_data_t* data = (ufo_seq_data_t*) userData;
    
    for (size_t i = 0; i < endValueIdx - startValueIdx; i++) {
        ((int *) target[i]) = data->from + (data->by - 1) * (i + startValueIdx);
    }

    return 0;
}
```

After we have written the function, all that is left is to plug it into our
`source` structure:

```C
SEXP/*INTXP*/ ufo_seq(SEXP/*INTXP*/ from, SEXP/*INTXP*/ to, SEXP/*INTXP*/ by) {

    ufo_source_t* source = (ufo_source_t*) malloc(sizeof(ufo_source_t));

    source->vector_type = UFO_INT; 
    source->element_size = sizeof(int);

    int from_value = INTEGER_ELT(from, 0);
    int to_value = INTEGER_ELT(to, 0);
    int by_value = INTEGER_ELT(by, 0);
    source->vector_size = (to_value - from_value) / by_value
                        + ((to_value - from_value) % by_value > 0);

    source->dimensions = NULL;
    source->dimensions_length = 0;

    source->min_load_count = 1000000 / sizeof(int);

    ufo_seq_data_t *data = (ufo_seq_data_t*) malloc(sizeof(ufo_seq_data_t));
    data->from = from_value;
    data->to = to_value;
    data->by = by_value;
    source->data = (ufUserData) data;

    source->destructor_function = &destroy_data;
    source->population_function = &populate;    

    ufo_new_t ufo_new = (ufo_new_t) R_GetCCallable("ufos", "ufo_new");
    return ufo_new(source);
}
```

## Et voilà

Now all the elements are in place. We can test our new vectors in R:

```R
library(ufoseq)
v <- ufo_seq(1, 100, 3)
v[1]
```
